---
title: "RcensusPkg"
author: Rick Dean
date: last-modified
format: 
  gfm: default
editor: source
---

The goal of RcensusPkg is to provide easy access to the US Census Bureau’s [datasets](https://www.census.gov/data/developers/data-sets.html) and collection of TIGER/Line Shapefiles providing plot geometries for states, counties, roads, landmarks, water, enumeration tracts/blocks for the entire United States. The only requirement is for the user to apply for and obtain a free access key issued from the Bureau. See [Guidance for Developers](https://www.census.gov/data/developers/guidance.html) for additional information.

Three functions (`get_vintage_data()`, `multi_vintage_data()`, `get_idb-data()`) call for a key along with many of the examples and tests. Be sure to read the description of these functions to learn more about incorporating the key.

The example below illustrates a simple workflow for downloading a dataset, merging the data with shapefile geometries, and plotting the merge to create a choropleth map. A more detailed example of the `RcensusPkg` workflow is available [here](https://rcensuspkg-ex.netlify.app/).

## Installation

The package is available for installation from [CRAN](https://cran.r-project.org/web/packages/available_packages_by_name.html).

You can install the development version of RcensusPkg from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("deandevl/RcensusPkg")
```

Using `devtools::install_github()`:

`devtools::install_github("deandevl/RcensusPkg")`

## Example

### Setup

We will be using the following packages.

```{r}
#| message: false
#| warning: false

library(httr)
library(jsonlite)
library(stringr)
library(data.table)
library(withr)
library(sf)
library(kableExtra)
library(ggplot2)
library(RplotterPkg)
library(RcensusPkg)
```

### A look at the US Census Bureau’s community resilience estimates (CRE) database.

Among the list of [available API](https://www.census.gov/data/developers/data-sets.html), there is the Community Resilience Estimates based on such factors as:

-   Income-to-Poverty Ratio (IPR) \< 130 percent
-   Single or zero caregiver household
-   Aged 65 years or older
-   No health insurance coverage
-   No vehicle access (Household)
-   Disability, at least one serious constraint to significant life activity
-   No one in the household is employed full-time, year-round
-   Households without broadband internet access
-   Unit-level crowding with \>= 0.75 persons per room
-   No one in the household has received a high school diploma
-   No one in the household speaks English “very well”

The factors are used to estimate the number of people with:

-   0 risk factors (Low risk)
-   1-2 risk factors (Medium risk)
-   3 or more risk factors (High risk)

The workflow for using `RcensusPkg` is the following:

1.  Get a database name recognized by the API. Use `RcensusPkg::get_dataset_names()` filtering for a “resilience” title and vintage of 2022.

```{r}
datasets_dt <- RcensusPkg::get_dataset_names(
  vintage = 2022,
  filter_title_str = "resilience"
)
```

```{r}
#| echo: false
#| tbl-cap: "Table 1: 'resilience' datasets"

kableExtra::kable(datasets_dt$data)
```

The returned dataframe shows a dataset name for CRE 2022 as (surprise) "cre".

2.  Get the variable names available for the "cre" dataset.

```{r}
cre_var_names_dt <- RcensusPkg::get_variable_names(
  dataset = "cre",
  vintage = 2022
) |> 
  _[, .(name, label)]
```

```{r}
#| echo: false
#| tbl-cap: "Table 2: Variable names from the CRE dataset"

kableExtra::kable(cre_var_names_dt)
```

We are interested in the percentage of individuals with three or more vulnerabilities ("PRED3_PE")

3.  Get the regions available for the CRE dataset.

```{r}
cre_regions_dt <- RcensusPkg::get_geography(
  dataset = "cre",
  vintage = 2022
)
```

```{r}
#| echo: false
#| tbl-cap: "Table 3: Regions from the CRE dataset"

kableExtra::kable(cre_regions_dt)
```

So we can get CRE estimates from the entire US, state, county, and tract enumeration levels. We are interested in the counties for the state of Florida.

4.  Download the data:

```{r}
florida_fips <-  usmap::fips(("FL"))

florida_cre_dt <- RcensusPkg::get_vintage_data(
  dataset = "cre",
  vintage = 2022,
  vars = "PRED3_PE",
  region = "county",
  regionin = paste0("state:", florida_fips)
) |> 
  _[, PRED3_PE := as.numeric(PRED3_PE)] |> 
  data.table::setnames(old = "PRED3_PE", new = "CRE_GE_3")
```

```{r}
#| echo: false
#| tbl-cap: "Table 4: Florida county percent risk from the CRE dataset"

kableExtra::kable(florida_cre_dt[1:8,])
```

5.  Merge the CRE county data with the Florida county shapefile geographies from the Bureau.

```{r}
#| message: false

output_dir <- withr::local_tempdir()
if(!dir.exists(output_dir)){
  dir.create(output_dir)
}
express <- parse(text = paste0("STATEFP == ", '"', florida_fips, '"'))
cre_florida_sf <- RcensusPkg::tiger_counties_sf(
  output_dir = output_dir,
  vintage = 2022,
  general = TRUE,
  express = express,
  datafile = florida_cre_dt,
  datafile_key = "county"
)
```

6.  Plot the choropleth map of the percent of individuals with 3 or more risk factors.

```{r}
RplotterPkg::create_sf_plot(
  sf = cre_florida_sf,
  aes_fill = "CRE_GE_3",
  hide_x_tics = TRUE,
  hide_y_tics = TRUE,
  panel_color = "white",
  panel_border_color = "white",
  caption = "Percent 3 or more risk factors among Florida counties"
)
```
